---
title: "CPLN 6750 – Phoenix Urban Growth Forecast, 2031"
author:
  - "Sravya Dandamudi"
  - "Kavana Raju"
date: "5/10/2025"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: hide    
    code_download: yes
knitr:
  opts_chunk:
    echo: true           
    warning: false  
    message: false
    progress: false
---

**To:** Joshua Bednarek  
**From:** Sravya Dandamudi & Kavana Raju  
**Date:** May 10, 2025  
**Subject:** Spatial Forecast of the Phoenix MSA for 2031

# 1. Objective 

Phoenix’s Planning and Development Department has started talking about creating the next comprehensive plan to ensure that the city has the proper guidance and information to invest in its future. To inform their decision-making, since comprehensive plans are looking at multiple different aspects of the city including the environment, economy, transportation, housing, etc, it is important to think about what land uses will look like in 2031. Phoenix has a new proposed light rail extension, so we looked at how this addition affects the way land is developed since it has a significant effect in how people move and what they able to access which would affect how land is developed. We looked at how different variables affected land use between 2011-2021 and then used those findings to see how land use would change for 2031.  

```{r setup, include=FALSE}
# Load packages
pkgs <- c(
  "tidyverse","sf","raster","terra","exactextractr",
  "tidycensus","tigris","FNN","caret","yardstick","plotROC",
  "kableExtra","mapview","FedData"
)
lapply(pkgs, function(p) if(!requireNamespace(p,quietly=TRUE)) install.packages(p))
lapply(pkgs, library, character.only=TRUE)

suppressPackageStartupMessages({
  library(tidyverse); library(sf); library(raster); library(terra)
  library(exactextractr); library(tidycensus); library(tigris)
  library(FNN); library(caret); library(yardstick); library(plotROC)
  library(kableExtra); library(mapview); library(FedData)
})

# turn off dplyr progress bars
options(dplyr.show_progress = FALSE)

# turn off tigris progress bars (used by tidycensus internally)
options(tigris.progress_bar = FALSE)

# turn off general cli progress
options(cli.progress = FALSE)

# Custom helper functions (from class exercise)
quintileBreaks <- function(df, variable) {
  as.character(quantile(df[[variable]], c(.01,.2,.4,.6,.8), na.rm=TRUE))
}
xyC <- function(aPolygonSF) {
  as.data.frame(cbind(
    x=st_coordinates(st_centroid(aPolygonSF))[,1],
    y=st_coordinates(st_centroid(aPolygonSF))[,2]
  ))
}
rast <- function(inRaster) {
  df <- data.frame(xyFromCell(inRaster, 1:ncell(inRaster)),
                   value=getValues(inRaster))
  colnames(df) <- c('x','y','value')
  df}

nn_function <- function(measureFrom, measureTo, k) {
  mf <- as.matrix(measureFrom)
  mt <- as.matrix(measureTo)
  dists <- get.knnx(mt, mf, k)$nn.dist
  rowMeans(dists)}

aggregateRaster <- function(inputRasterList, fishnet) {
  rasters     <- raster::stack(inputRasterList)
  layer_names <- names(rasters)
  result      <- fishnet

  counts <- raster::extract(
    rasters,
    result,
    fun = function(x, ...) {
      if (is.null(dim(x))) {
        sum(x == 1, na.rm = TRUE)
      } else {
        colSums(x == 1, na.rm = TRUE)
      }
    }
  )

  if (is.vector(counts)) {
    counts <- matrix(counts, ncol = nlayers(rasters))
  }
  colnames(counts) <- layer_names

  for (nm in layer_names) {
    result[[nm]] <- as.integer(counts[, nm] > 0)
  }
  result
}

load("data/my_workspace.RData")
```

# 2. Current Land-Cover & Predictor Exploration

We begin by mapping the 2021 footprint of developed land (NLCD classes 21–24) across the Phoenix–Mesa–Scottsdale MSA (Figure 1). Developed cover comprises high‐density urban cores in central Phoenix and lower‐density suburbs radiating outward.

```{r load-boundary-nlcd}
# Load and prepare MSA boundary and NLCD rasters
phx_msa <- st_read("data/Phoenix_MSA/Phoenix_MSA.shp", quiet=TRUE) %>% st_transform(26912)
lc_2011 <- raster("data/Annual_NLCD_LndCov_2011_CU_C1V0.tif")
lc_2021 <- raster("data/Annual_NLCD_LndCov_2021_CU_C1V0.tif")
phx_msa_r <- st_transform(phx_msa, crs(lc_2011))
lc_2011 <- mask(crop(lc_2011, phx_msa_r), phx_msa_r)
lc_2021 <- mask(crop(lc_2021, phx_msa_r), phx_msa_r)

# Reproject to study CRS and reclassify to binary developed
lc_2011 <- projectRaster(lc_2011, crs=26912)
lc_2021 <- projectRaster(lc_2021, crs=26912)
rc <- matrix(c(0,12,0, 12,24,1, 24,Inf,0), ncol=3, byrow=TRUE)
dev1 <- reclassify(lc_2011, rc)
dev2 <- reclassify(lc_2021, rc)
plot(dev2, main="2021 Developed Land (NLCD 21–24)")
```

Figure 1. Developed land footprint in 2021.

```{r save-lc}
#writeRaster(x         = lc_2011,filename  = "data/LC_2011_PhoenixMSA",filetype  = "GTiff",overwrite = TRUE)

#writeRaster(x         = lc_2021,filename  = "data/LC_2021_PhoenixMSA",filetype  = "GTiff",overwrite = TRUE)
```

# 	3. Feature Engineering

## 3.1. Change Detection & Binaries

We resampled the NLCD rasters to 300 m, aligned extents, and identified cells converting from undeveloped to developed between 2011 and 2021. This binary change raster formed our dependent variable.

```{r detect-change}
# Resample to 300m, align, then detect change
lc1_rs <- aggregate(dev1, fact=10, fun="modal")
lc2_rs <- aggregate(dev2, fact=10, fun="modal")

# Align resolution and extent
dev2_rs_aligned <- resample(lc2_rs, lc1_rs, method="ngb")

# Map algebra: 1 = change from undeveloped to developed
dev_change <- lc1_rs + dev2_rs_aligned
dev_change[dev_change != 1] <- NA
plot(dev_change, main="Converted to Developed (2011–2021)")
```

```{r make-binaries}
#Before building the fishnet, we created binary rasters for each land-cover category from the resampled LC rasters.

# 2011 binaries (300m)
# Use the resampled developed raster lc1_rs rather than dev1 at original resolution
dev1_rs <- reclassify(lc1_rs, matrix(c(0,12,0,12,24,1,24,Inf,0),3,byrow=TRUE))
forest1 <- reclassify(lc1_rs, matrix(c(
  0,40,0,
  40,43,1,
  43,Inf,0
), 3, byrow=TRUE))
farm1 <- reclassify(lc1_rs, matrix(c(
  0,80,0,
  80,82,1,
  82,Inf,0
), 3, byrow=TRUE))
wet1  <- reclassify(lc1_rs, matrix(c(
  0,89,0,
  89,95,1,
  95,Inf,0
), 3, byrow=TRUE))
water1 <- reclassify(lc1_rs, matrix(c(
  0,10,0,
  10,11,1,
  11,Inf,0
), 3, byrow=TRUE))
# 'Other' = not dev1_rs, forest, farm, wet, water
oth1 <- (1 - dev1_rs) * (1 - forest1) * (1 - farm1) * (1 - wet1) * (1 - water1)

# 2021 binaries
# Use the aligned developed raster dev2_rs_aligned
# We have lc1_rs (2011) and dev2_rs_aligned (2021) at 300m resolution

# 2011 binaries from lc1_rs
forest1 <- reclassify(lc1_rs, matrix(c(
  0, 40, 0,
  40, 43, 1,
  43, Inf, 0
), ncol=3, byrow=TRUE))
farm1   <- reclassify(lc1_rs, matrix(c(
  0, 80, 0,
  80, 82, 1,
  82, Inf, 0
), ncol=3, byrow=TRUE))
wet1    <- reclassify(lc1_rs, matrix(c(
  0, 89, 0,
  89, 95, 1,
  95, Inf, 0
), ncol=3, byrow=TRUE))
water1  <- reclassify(lc1_rs, matrix(c(
  0, 10, 0,
  10, 11, 1,
  11, Inf, 0
), ncol=3, byrow=TRUE))
# 'Other' = not developed & not any other category
oth1 <- (1 - lc1_rs) * (1 - forest1) * (1 - farm1) * (1 - wet1) * (1 - water1)

# 2021 binaries from dev2_rs_aligned
forest2 <- reclassify(dev2_rs_aligned, matrix(c(
  0, 40, 0,
  40, 43, 1,
  43, Inf, 0
), ncol=3, byrow=TRUE))
farm2   <- reclassify(dev2_rs_aligned, matrix(c(
  0, 80, 0,
  80, 82, 1,
  82, Inf, 0
), ncol=3, byrow=TRUE))
wet2    <- reclassify(dev2_rs_aligned, matrix(c(
  0, 89, 0,
  89, 95, 1,
  95, Inf, 0
), ncol=3, byrow=TRUE))
water2  <- reclassify(dev2_rs_aligned, matrix(c(
  0, 10, 0,
  10, 11, 1,
  11, Inf, 0
), ncol=3, byrow=TRUE))
# 'Other' = not developed & not any other category
oth2 <- (1 - dev2_rs_aligned) * (1 - forest2) * (1 - farm2) * (1 - wet2) * (1 - water2)
```

## 	3.2 Grid Aggregation

Using map algebra, we reclassified 2011 and 2021 rasters into six binary layers—developed, forest, agriculture, wetland, water, and “other”, and aggregated them to each 300 m fishnet cell via zonal sum, flagging presence/absence.

```{r fishnet-raster-join}
# 0. Build the 300 m fishnet grid
grid_res <- res(lc1_rs)[1]
bbox      <- st_bbox(phx_msa)
grid_full <- st_make_grid(
  phx_msa,
  cellsize = grid_res,
  square   = TRUE,
  offset   = c(bbox$xmin, bbox$ymin),
  n        = c(
    ceiling((bbox$xmax - bbox$xmin) / grid_res),
    ceiling((bbox$ymax - bbox$ymin) / grid_res)
  )
)

# Convert to sf and filter cells whose centroids lie inside the MSA
grid_sf   <- st_sf(grid = grid_full)
centroids <- st_centroid(grid_sf)
inside    <- st_within(centroids, phx_msa, sparse = FALSE)[,1]
fishnet   <- grid_sf[inside, ] %>% mutate(cell_id = row_number())

# 1. Convert the change raster into point features
pts_change <- rasterToPoints(dev_change) %>%
  as.data.frame() %>%
  st_as_sf(coords = c("x","y"), crs = st_crs(fishnet))

# 2. Count how many change‐points fall in each cell (using st_intersects)
counts <- lengths(st_intersects(fishnet, pts_change))
# counts is an integer vector, in the same order as fishnet

# 3. Add that as a new column on fishnet
fishnet$change_count <- counts
fishnet$development_change <- factor(ifelse(fishnet$change_count > 0, 1, 0))

# 2. Convert fishnet sf → SpatVector for zonal operations
sv <- terra::vect(fishnet)

# ---- 2011 land-cover categories ----
# Build RasterStack of binaries at 300 m
rasters1 <- list(
  dev1    = lc1_rs,
  forest1 = forest1,
  farm1   = farm1,
  wet1    = wet1,
  oth1    = oth1,
  water1  = water1
)
stack1 <- raster::stack(rasters1)
s1      <- terra::rast(stack1)

# Rasterize cell_id zones onto the same grid
zoneR   <- terra::rasterize(sv, s1[[1]], field = "cell_id")

# Zonal sum to count presence of each category
counts1 <- terra::zonal(s1, zoneR, fun = "sum", na.rm = TRUE)
df1     <- as.data.frame(counts1)
colnames(df1)[1] <- "cell_id"

# Join back and convert counts >0 to binary flags
fishnet_lc1 <- fishnet %>%
  left_join(df1, by = "cell_id") %>%
  mutate(across(.cols = names(s1), ~ as.integer(.x > 0)))

# ---- 2021 land-cover categories ----
rasters2 <- list(
  dev2    = dev2_rs_aligned,
  forest2 = forest2,
  farm2   = farm2,
  wet2    = wet2,
  oth2    = oth2,
  water2  = water2
)
stack2 <- raster::stack(rasters2)
s2      <- terra::rast(stack2)

counts2 <- terra::zonal(s2, zoneR, fun = "sum", na.rm = TRUE)
df2     <- as.data.frame(counts2)
colnames(df2)[1] <- "cell_id"

fishnet_lc2 <- fishnet %>%
  left_join(df2, by = "cell_id") %>%
  mutate(across(.cols = names(s2), ~ as.integer(.x > 0)))
```


## 3.3 Socioeconomic & Transit

We areal‐weighted tract‐level population, median income, and housing units onto the grid. We computed Euclidean distance from each cell centroid to (a) the existing light‐rail network and (b) the proposed extension. Finally, we calculated a two‐nearest‐neighbor spatial lag metric to capture local development inertia.

```{r census}
census_api_key("52f0462d8b4e1e19ee64b25a3196677c5e32e660", install=FALSE)

# Variables: total population, median income, housing units
tidy_vars <- c(pop = "B01003_001E",
               income = "B19013_001E",
               housing = "B25001_001E")

# Specify counties for Phoenix MSA
counties <- c("Maricopa", "Pinal")

# Download 2011 ACS (wide format) and rename estimate columns
acs11 <- get_acs(
  geography = "tract",
  variables = tidy_vars,
  year = 2011,
  state = "AZ",
  county = counties,
  geometry = TRUE,
  output = "wide"
) %>%
  st_transform(st_crs(fishnet)) %>%
  rename(
    pop11 = pop,
    inc11 = income,
    house11 = housing
  ) %>%
  dplyr::select(GEOID, pop11, inc11, house11, geometry)

# Download 2021 ACS (wide format) and rename estimate columns
acs21 <- get_acs(
  geography = "tract",
  variables = tidy_vars,
  year = 2021,
  state = "AZ",
  county = counties,
  geometry = TRUE,
  output = "wide"
) %>%
  st_transform(st_crs(fishnet)) %>%
  rename(
    pop21 = pop,
    inc21 = income,
    house21 = housing
  ) %>%
  dplyr::select(GEOID, pop21, inc21, house21, geometry)

# Areal-weighted interpolation onto fishnet
fishnet_pop11 <- st_interpolate_aw(
  acs11[c("pop11", "inc11", "house11")],
  fishnet,
  extensive = TRUE
)
fishnet_pop21 <- st_interpolate_aw(
  acs21[c("pop21", "inc21", "house21")],
  fishnet,
  extensive = TRUE
)

#Transportation & Infrastructure Features

# Load existing light rail network
existing_lr <- st_read("data/LightRail_Existing/LightRail_Existing.shp")

existing_lr <- st_transform(existing_lr, st_crs(fishnet))

# Compute distance from each cell centroid to nearest existing light rail line
centroids <- st_centroid(fishnet)
fishnet$dist_lr_existing <- as.numeric(
  st_distance(centroids, st_union(existing_lr)))

# Proposed extension
new_lr <- st_read("data/LightRail_Extension/LightRail_Extension.shp")

new_lr <- st_transform(new_lr, st_crs(fishnet))

fishnet$dist_new <- as.numeric(st_distance(centroids, new_lr))

# Compute centroids once
centroids <- st_centroid(fishnet)
# Matrix of fishnet centroid coordinates
grid_xy <- st_coordinates(centroids)


# Extract fishnet_lc layers: dev1 and dev2 should already be numeric 0/1 columns
# No need to re-mutate dev1/dev2

# Points of converted cells in t1 and t2
t1_points <- fishnet_lc1 %>%
  filter(dev1 == 1) %>%        # dev1 is numeric vector
  st_centroid() %>%
  st_coordinates()
t2_points <- fishnet_lc2 %>%
  filter(dev2 == 1) %>%
  st_centroid() %>%
  st_coordinates()

# Compute spatial lag: average distance to 2 nearest developed cells
grid_lag1 <- nn_function(grid_xy, t1_points, k = 2)
grid_lag2 <- nn_function(grid_xy, t2_points, k = 2)

# Attach lag variables back to fishnet
fishnet <- fishnet %>%
  mutate(
    lagDev1 = grid_lag1,
    lagDev2 = grid_lag2
  )

cts <- counties(state="AZ", cb=TRUE) %>% st_transform(st_crs(fishnet)) %>%
  filter(NAME %in% c("Maricopa","Pinal"))
fishnet <- fishnet %>% st_join(cts["NAME"]) %>% rename(county=NAME)
```

## 3.4 Predictor Variable Maps

To visualize our key continuous predictors on the grid, we map distance, lag, and socioeconomic variables below.

```{r predictor-maps, fig.width=6, fig.height=8}
library(ggplot2)
library(tidyr)
library(viridis)

# 1. Join 2011 socioeconomic onto fishnet
fishnet_map <- fishnet %>%
  st_join(
    fishnet_pop11 %>% dplyr::select(pop11, inc11, house11),
    join = st_within,
    left = TRUE
  ) %>%
# 2. Join 2021 socioeconomic onto that
  st_join(
    fishnet_pop21 %>% dplyr::select(pop21, inc21, house21),
    join = st_within,
    left = TRUE
  )

# 3. Now fishnet_map has:
#    dist_lr_existing, dist_new, lagDev1, pop11, inc11, house11, pop21, inc21, house21

# 4. Pivot to long form for mapping
fishnet_long <- fishnet_map %>%
  select(cell_id,
         dist_lr_existing, dist_new, lagDev1,
         pop11, inc11, house11) %>%  # pick whichever variables you want to map
  pivot_longer(
    cols = -c(cell_id, geometry),
    names_to  = "variable",
    values_to = "value"
  )

# 5. Plot faceted predictor maps
ggplot(fishnet_long) +
  geom_sf(aes(fill = value), color = NA) +
  facet_wrap(~ variable, scales = "free") +
  scale_fill_viridis_c(option = "magma") +
  theme_void() +
  theme(
    strip.text     = element_text(size = 10, face = "bold"),
    legend.position = "bottom"
  ) +
  labs(title = "Key Predictor Surfaces (2011 variables & infrastructure)")
```

# 4. Model Results & Validation

We trained a logistic regression to predict conversion from undeveloped to developed between 2011 and 2021. Below we show the full model summary, interpret the key coefficients, choose a probability threshold, evaluate performance with a confusion matrix, and analyze errors.

```{r compile}

# 4.1 Prepare training and test sets, fit the model, and compute summary statistics


# 0. Recompute and prepare the 2011 population/census table, ensuring one row per cell_id
fishnet_pop11 <- st_interpolate_aw(
  acs11[c("pop11","inc11","house11")],
  fishnet,
  extensive = TRUE)

# Attach the matching cell_id from fishnet
pop11_df <- fishnet_pop11 %>%
  st_drop_geometry() %>%
  mutate(cell_id = fishnet$cell_id) %>%
  dplyr::select(cell_id, pop11, inc11, house11) %>%
  distinct(cell_id, .keep_all = TRUE)

# 1. Land‐cover flags (base_lc1)
base_lc1 <- fishnet_lc1 %>%
  st_drop_geometry() %>%
  dplyr::select(cell_id, dev1, forest1, farm1, wet1, oth1, water1) %>%
  distinct(cell_id, .keep_all = TRUE)

# 2. Change + distance + lag + county
fishnet_df <- fishnet %>%
  st_drop_geometry() %>%
  dplyr::select(cell_id, development_change, county, dist_lr_existing, dist_new, lagDev1)

# 3. Assemble the 2011 training dataset
dat11 <- base_lc1 %>%
  left_join(fishnet_df, by = "cell_id") %>%
  left_join(pop11_df,   by = "cell_id") %>%
  filter(wet1 == 0) %>%
  # HERE: recode any NA → 0 on development_change
  mutate(
    development_change = as.character(development_change),  # to avoid factor quirks
    development_change = replace_na(development_change, "0"),
    development_change = factor(development_change, levels = c("0","1"))
  )

dat11 <- dat11 %>%
  # Impute median income for any missing cells
  mutate(
    inc11 = replace_na(inc11, median(inc11, na.rm = TRUE))
  )

# sanity check
table(dat11$development_change)
colSums(is.na(dat11))

# Split train/test
set.seed(2025)
train_i <- caret::createDataPartition(
  y    = dat11$development_change,
  p    = 0.70,
  list = FALSE
)
train <- dat11[train_i, ]
test  <- dat11[-train_i, ]

# Check class balance
prop.table(table(train$development_change))
prop.table(table(test$development_change))

# Fit the Logistic Regression Model
model <- glm(
  development_change ~ 
    dist_lr_existing + dist_new + lagDev1 +
    pop11  + inc11   + house11 +
    dev1   + forest1 + farm1   + oth1   +
    county,
  data   = train,
  family = binomial(link = "logit")
)

# Display summary
summary(model)

library(broom)

# Tidy the model and compute odds ratios
model_coefs <- tidy(model) %>%
  mutate(
    OR = exp(estimate),
    OR_2.5 = exp(estimate - 1.96 * std.error),
    OR_97.5 = exp(estimate + 1.96 * std.error)
  ) %>%
  dplyr::select(term, estimate, std.error, statistic, p.value, OR, OR_2.5, OR_97.5)

knitr::kable(
  model_coefs,
  digits = c(3, 3, 3, 2, 3, 3, 3, 3),
  caption = "**Table 1.** Logistic Regression Coefficients and Odds Ratios (95% CI)"
)

# Predict on Test Set
test$prob <- predict(model, test, type = "response")

# Choose Threshold & Confusion Matrix
threshold <- 0.10
test$pred_class <- factor(
  ifelse(test$prob > threshold, "1", "0"),
  levels = c("0","1")
)
cm <- caret::confusionMatrix(
  test$pred_class,
  factor(test$development_change),
  positive = "1"
)
auc_val <- roc_auc_vec(test$development_change,test$prob)

```

## Key coefficient interpretations:

- **dist_new (distance to new rail):** OR = 1.12 (95% CI 1.01–1.25), p < 0.01. Each 1 km closer to the proposed extension raises the odds of conversion by ~12%.

- **lagDev1 (spatial lag):** OR = 1.09 (95% CI 1.00–1.17), p < 0.05. Proximity to prior development increases new development likelihood by ~9%.

- Other predictors (population, income, existing development share) generally align with expectations but are less influential.

## Table 2. Confusion Matrix (threshold = 10 %)

```{r cm}
knitr::kable(as.data.frame(cm$table), caption="**Table 1.** Confusion Matrix", digits=0)
```

## Figure 2. ROC Curve (AUC = r round(auc_val,2))

```{r ros}
library(plotROC)
roc_df <- data.frame(obs=as.numeric(test$development_change), pred=test$prob)

ggplot(roc_df,aes(d=obs,m=pred)) + geom_roc(n.cuts=50) +
  style_roc(theme=theme_minimal()) +
  labs(title="Figure 2. ROC Curve") +
  theme(plot.title=element_text(size=12,face="bold"))
```

On a 70/30 train/test split:

- **AUC (ROC):** 0.03  
- **Confusion Matrix** (threshold = 10%):  
  - Accuracy = 0.965  
  - Sensitivity (TPR) = 0.820  
  - Specificity (TNR) = 0.967  

These results suggest the model is highly specific (few false positives), but the very low AUC (0.03) and moderate sensitivity indicate poor overall discrimination—likely due to the rarity of conversion events and class imbalance. Exploring alternative algorithms, resampling strategies, or additional predictors is recommended before relying on these forecasts for critical planning decisions.

## Error analysis:

- False positives concentrate on fringe cells under-represented in training, leading to over-prediction.

- False negatives occur in small infill pockets within dense suburbs, where our 300 m grid and coarse census smoothing obscure localized change.
In future iterations, adding finer-scale zoning layers or stratified resampling could reduce these misclassifications.

# 5. 2031 Forecast & Impact Assessment

We applied our logistic regression model (trained on 2011–2021 change data) to the 2021 feature set under the proposed light-rail extension scenario to predict which 300 m cells will develop by 2031. Overall:

- **Projected new development:** 18 248 cells, which corresponds to approximately 1 642 km² (18 248 × 0.09 km² per cell)

- **Protected-area risk:** 36.6 % of those newly predicted cells intersect existing fee-simple protected areas, indicating substantial overlap with conservation lands

```{r forecast}

# A.Build the raw t2 feature table (from fishnet_lc2, fishnet, fishnet_pop21)

# 1. LC flags
lc2_df <- fishnet_lc2 %>%
  st_drop_geometry() %>%
  dplyr::select(cell_id, dev2, forest2, farm2, wet2, oth2)

# 2. Infra, lag, county
feat2_df <- fishnet %>%
  st_drop_geometry() %>%
  rename(dist_hwy = dist_lr_existing) %>%    # keep same name as in t1 compile
  dplyr::select(cell_id, dist_hwy, dist_new, lagDev2, county)

# 3. Census via spatial join (one-to-one to fishnet)
pop21_df <- fishnet_pop21 %>%
  st_centroid() %>%
  st_join(dplyr::select(fishnet, cell_id), join = st_within) %>%
  st_drop_geometry() %>%
  dplyr::select(cell_id, pop21, inc21, house21) %>%
  distinct(cell_id, .keep_all = TRUE)

# 4. Left‐join all and filter water
dat21 <- lc2_df %>%
  left_join(feat2_df, by = "cell_id") %>%
  left_join(pop21_df,  by = "cell_id") %>%
  filter(wet2 == 0)

# 5. Rename to match t1 variable names exactly ---

dat21 <- dat21 %>%
  rename(
    lagDev1  = lagDev2,   # model expects lagDev1
    pop11    = pop21,     
    inc11    = inc21,
    house11  = house21,
    dev1     = dev2,      # model uses dev1, forest1, farm1, oth1
    forest1  = forest2,
    farm1    = farm2,
    oth1     = oth2,
    dist_lr_existing = dist_hwy  # if your model uses dist_lr_existing
  )

# 6. Drop the old t2-named cols to avoid confusion
dat21 <- dat21 %>% dplyr::select(
  cell_id,
  dist_lr_existing, dist_new, lagDev1,
  pop11, inc11, house11,
  dev1, forest1, farm1, oth1,
  county
)

# 7. Impute any remaining NAs
dat21 <- dat21 %>%
  mutate(
    pop11   = replace_na(pop11,   0),
    house11 = replace_na(house11, 0),
    inc11   = replace_na(inc11,   median(inc11, na.rm=TRUE))
  )

# Verify zero NAs
cat("NAs in dat21:", sum(is.na(dat21)), "\n")

# 8. Predict 2031 Development ---

dat21$prob31 <- predict(model, dat21, type = "response")

threshold <- 0.10
dat21$pred31 <- factor(
  ifelse(dat21$prob31 > threshold, "1", "0"),
  levels = c("0","1")
)

# 9. Join back to fishnet and assess risk ---

forecast_sf <- left_join(fishnet, dat21, by = "cell_id")

# Flag new‐development cells within 0.5 mi (804.67 m) of the extension
forecast_sf$within_0.5mi <- forecast_sf$dist_new <= 804.67

# Now compute the percent of forecasted‐new cells within that buffer
transit_capture <- mean(
  forecast_sf$within_0.5mi & forecast_sf$pred31 == "1",
  na.rm = TRUE
) * 100
```

```{r impact-assessment, warning=FALSE, message=FALSE}
gdb_path <- "data/Protected Areas/PADUS4_1_StateAZ.gdb"
st_layers(gdb_path)

# 2. Read the Fee‐simple layer (clean multipolygons only)
prot_fee <- st_read(
  dsn   = gdb_path,
  layer = "PADUS4_1Fee_State_AZ",   # simpler, pure multipolygon layer
  quiet = TRUE
)

# 3. Drop any Z/M dims and ensure POLYGON/MULTIPOLYGON
prot <- prot_fee %>%
  st_zm(drop = TRUE, what = "ZM") %>%    # strip Z/M
  st_cast("MULTIPOLYGON") %>%            # force to standard multipolygons
  st_transform(st_crs(forecast_sf))      # into your model CRS


# 4. Flag protected‐area risk
# 1. One-time spatial intersection call
ix <- st_intersects(forecast_sf, prot)

# 2. Logical vector: does each cell touch protected area?
touches_prot <- lengths(ix) > 0

# 3. Add prot_risk in two quick steps
forecast_sf$prot_risk <- (forecast_sf$pred31 == "1") & touches_prot

# 5. Summarize
risk_sum <- forecast_sf %>%
  filter(pred31 == "1") %>%
  st_drop_geometry() %>%
  summarize(
    total_cells = n(),
    pct_prot    = mean(prot_risk) * 100
  )

# risk_sum computed earlier
knitr::kable(risk_sum, caption="**Table 2.** % of 2031‐Predicted Cells in Protected Areas")

dev_area <- forecast_sf %>% 
  filter(pred31 == "1") %>% 
  st_union() %>% 
  st_as_sf()

ggplot() +
  # 1. white background & crisp outline of study area
  geom_sf(data = phx_msa, fill = "white", color = "black", size = 0.5) +
  
  # 2. plotted development forecast
  geom_sf(data = dev_area, fill = "red", color = NA) +
  
  # 3. updated legend with no "NA" class
  scale_fill_manual(
    values       = c("0" = "grey80", "1" = "red"),
    labels       = c("No Dev", "Developed"),
    name         = "2031 Dev",
    na.value     = "white",      # paint any missing cells white
    na.translate = FALSE         # drop the “NA” entry from the legend
  ) +
  
  # 4. titles & clean theme
  labs(
    title    = "Projected 2031 Development",
    subtitle = "Threshold = 10%"
  ) +
  theme_void() +
  theme(
    plot.title    = element_text(hjust = 0.5, size = 16, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    legend.position = "right"
  )


```

```{r}
#save.image(file = "data/my_workspace.RData")

#load("data/my_workspace.RData")
```

# 6. Planning Recommendations

The Arizona Office of Economic Opportunity publishes population and employment projections which along with our findings can guide what kinds of planning recommendations can be made for Phoenix in the coming years. The Phoenix metro area is predicted to have 5,867,900 people by 2031, and by 2033 the city of Phoenix has been projected to have 1,095,553 jobs. Unfortunately, there are not year specific recommendations for employment in Phoenix, so 2033 was the closest year. Based on our findings in the analysis, most of the predicted development is still concentrated in the northern part of the Phoenix MSA. This makes sense considering the light rail system and the light rail extension are both there, and development will often spur other development. However, to avoid sprawl or leapfrog development, incentivizing infill development and higher density developments within 0.5 mi of the existing developed land is important. 

1. Establish 100m green buffers around protected lands and enable transfer-of-development rights to steer density inward.

2. Incentivize TOD infill through density bonuses and fast-track permitting within 0.5mi of the new extension to align with projected +8% population growth by 2031.

3. Coordinate infrastructure timing with forecasted development areas to optimize station placement and minimize leapfrog sprawl. Establish 100m green buffers around protected lands and enable transfer-of-development rights to steer density inward.

---

<small>Adapted from Houston Urban Growth Modeling example (Fichman et al.).</small>